---
title: | 
  | Antibody Responses to Different Proteins 
  | in Prostate Cancer Patients  
author: "Tun Lee Ng and Michael A. Newton"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document:
    toc: true
    number_sections: true
header-includes:
   \usepackage{enumerate}
   \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

# Introduction

This project aims to characterize antibody responses to a wide variety of proteins in prostate cancer patients at different stages of the disease. 16-mer peptides spanning the amino acid sequences of these 1611 gene products, and overlapping by 12 amino acids, were used to generate a microarray comprising 177,604 peptides. In this study, there were healthy subjects and patients with different stages of prostate cancer  

*  ``new_dx``: newly diagnosed,
* ``nmCSPC``: non-metastatic castration-sensitive
* ``mCSPC``: metastatic castration-sensitive,
* ``nmCRPC``: non-metastatic castration-resistant,
* ``mCRPC``: metastatic castration-resistant  

Each patient’s serum was assayed in a number of replicates, ``rep``, which were 1, 2, or 3, for peptide-specific IgG responses using the microarray. We remove patients with ``rep`` = 1. The criterion for a positive call on a peptide for a patient was that they had to meet the signal (fluorescence level) threshold in at least two of the technical replicates. Since this is not possible for patients without technical replicates, we exclude them for consistency.

```{r}
library(allez)
library(kableExtra)
library(ggplot2)
library(lme4) # linear mixed effects model
library(fdrtool)
library(Rtsne)
library(heatmap3)
library(tidyverse) # make sure you have the latest tidyverse !

####################################################################################### 
#                           Some Common Variables and Functions                       #
####################################################################################### 

# specified color and shape scheme 
pal <- c("navy", "cornflowerblue", "turquoise1", "orchid1", "darkorange1", "firebrick1")
names(pal) <- c("normal",  "new_dx", "nmCSPC", "mCSPC", "nmCRPC", "mCRPC")
shp <- c(8, 15, 16, 3, 17, 18)
names(shp) <- names(pal)


# function to plot PCA loadings
PCload.func <- function(cols, shapes, U, D, x, y, pca.vec, title){
  Z <- U %*% diag(D)
  plot(Z[,x], Z[,y], col = cols, pch = shapes, main = title,
       xlab = paste0("PC",x," (", round(pca.vec[x]*100, 1) ,"% variance explained)"),
       ylab = paste0("PC",y," (", round(pca.vec[y]*100, 1) ,"% variance explained)") 
  )
}


# function to count peptides at different FDR thresholds
count.func <- function(pval.vec, thresh.vec){
  counter <- NULL
  for (i in thresh.vec ){
    counter <- c(counter, length(pval.vec[pval.vec <= i]))
  }
  countab <- rbind(c("FDR threshold", thresh.vec),
                   c("Peptide counts", counter))
  return(countab)
}


# typical step in ANOVA
anova_func <- function(anova_pval){
  # control FDR
  anova_BH <- p.adjust(anova_pval, method = "BH")
  anova_qval <- fdrtool(anova_pval, statistic = "pvalue", verbose = F, plot  = F)$qval
  anova_qval_eta0 <- unname(fdrtool(anova_pval, statistic = "pvalue", verbose = F, plot  = F)$param[,"eta0"])
  
  # plot histogram of p-values
  all_peptide_hist <- hist(anova_pval, breaks = 70, freq = F, xlab = "one-way ANOVA p-values", 
                           main = paste0("p-values distribution for ", length(anova_pval), " peptides"))
  polygon_ind <- which(all_peptide_hist$density >= anova_qval_eta0)
  for (i in polygon_ind){
    polygon( x = c(all_peptide_hist$breaks[i], all_peptide_hist$breaks[i+1], all_peptide_hist$breaks[i+1], all_peptide_hist$breaks[i]),
             y = c(all_peptide_hist$density[i], all_peptide_hist$density[i], anova_qval_eta0, anova_qval_eta0),
             col = "red")
  }
  text(x=0.65,y=4, labels = paste( "estimated proportion of \nnon-null peptides =",
                                   round( 100*(1 - anova_qval_eta0),2 ),"%" ))
  
  return(list(anova_BH = anova_BH, anova_qval = anova_qval, anova_qval_eta0 = anova_qval_eta0))
}


# commonly-used ggplot theme 
myTheme <- theme(panel.background = element_rect(fill = "grey90"),
                 panel.grid.major = element_line(color = "white"),
                 panel.grid.minor = element_line(color = "white"),
                 axis.text.x = element_text(angle = 90, hjust = 1, size = 4.5),
                 legend.position = "bottom",
                 plot.title = element_text(hjust = 0.5))


load("04_aggregated_calls.RData")
raw_data_median <- read_csv("raw_data_median.csv")

####################################################################################### 
#                                      Data Processing                                #
####################################################################################### 

array_id_key = read_csv("sample_key_project1.csv") %>%
  janitor::clean_names() %>% 
  rename(stage = condition,
         array_id = "file_name") %>%
  mutate(id = str_replace_all(id, " ", ""),
         id = str_to_lower(id),
         id = str_replace_all(id, "/", ""),
         stage = as_factor(stage),
         stage = fct_recode(stage,
                            "normal" = "Normal_male_controls",
                            "new_dx" = "Newly_diagnosed",
                            "nmCSPC" = "PSA-recurrent_nonMet",
                            "mCSPC" = "Met",
                            "nmCRPC" = "Castration-resistent_nonMet",
                            "mCRPC" = "Castration-resistent_Met",
                            "binding_buffer" = "Binding buffer alone"))

# drop binding buffer
array_id_key <- array_id_key[!(array_id_key$stage=="binding_buffer"),]
array_id_key$stage <- factor(array_id_key$stage) # remove binding_buffer level

# remove patients whose rep == 1
patient_key = array_id_key %>%
  group_by(id, stage) %>%
  summarize(n = n()) %>%
  ungroup() %>% 
  filter(n >= 2) %>%
  select(-n)
array_id_key = array_id_key %>%
  filter(id %in% patient_key$id)

# there are patients who were measured at two different stages
# To ensure unique patients, remove the following ids

ids_to_remove = c("adt181",
                  "adt223",
                  "pdv008",
                  "pap123",
                  "pap067",
                  "adt143")

# drop patients' earlier records
array_id_key = array_id_key %>%
  filter(!(id %in% ids_to_remove))
patient_key = array_id_key %>%
  group_by(id, stage) %>%
  tally() %>%
  select(-n) 
patient_key$stage <- relevel(patient_key$stage, ref = "normal")
```

Note that these `r array_id_key %>% group_by(id, stage) %>% summarize() %>% group_by(stage) %>% tally() %>% pull(n) %>% sum()` patient-records do not represent distinct patients, as certain patients were measured at two different stages of prostate cancer. There are `r length(ids_to_remove)` patients who had replicates for both stages. We removed their earlier-stage records, and finally arrive at `r nrow(patient_key)` distinct patients. 

```{r, fig.height = 2.6}
distinct_patients_with_reps <- array_id_key %>%
  group_by(id, stage) %>%
  summarize() %>%
  group_by(stage) %>%
  tally()

ggplot(data=distinct_patients_with_reps, aes(x=stage, y=n)) + 
  geom_bar(stat="identity", width=0.5 , fill="steelblue") + 
  geom_text(aes(label=n), vjust=-0.3, size=2.5) +
  theme_minimal()
```

We will utilize both fluorescence levels data and patients' binary calls data to **investigate if patients at different stages of prostate cancer exhibit different antibody responses to certain peptide chains or proteins**. We take $\log_2$ transformation on the fluorescence levels prior to subsequent steps in our analysis. In order to verify normalization of the fluorescence level, we also plot the boxplots of median (across replicates) $\log_2$ fluorescence level of all peptides for each patient.

```{r, fig.height = 5, cache = T, dependson=c("myTheme", "raw_data_median", "patient_key", "array_id_key")}
####################################################################################### 
#                         Check normalization of fluorescence data                    #         
#######################################################################################

median_long <- raw_data_median %>%
  select(any_of(array_id_key$id)) %>%
  pivot_longer(cols = everything(), names_to = "id", values_to = "fluorescence") 

# set fill color
median_long$stage <- patient_key$stage[ match(median_long$id, patient_key$id) ]
# median_long$cols <- pal[ match(median_long$stage, names(pal)) ]

# sort order of patients in boxplot
median_long$id <- factor(median_long$id, levels = c(
  patient_key$id[patient_key$stage == "normal"],
  patient_key$id[patient_key$stage == "new_dx"],
  patient_key$id[patient_key$stage == "nmCSPC"],
  patient_key$id[patient_key$stage == "nmCRPC"],
  patient_key$id[patient_key$stage == "mCRPC"]
))

ggplot(median_long, aes(x = id, y = fluorescence, fill = stage)) +
  geom_boxplot() +
  # scale_fill_discrete(name="Stage") +
  # guides(fill=guide_legend(title="Stage")) +
  scale_fill_manual(name = "Stage", values = pal) +
  labs(title = "Boxplots of Peptide Fluorescence Levels for All Patients", 
       x = "Patient ID", y = "Median Fluorescence Levels on log2 scale") +
  ylim(2,16) +
  myTheme
```

It appears that the fluorescence levels of the peptides are normalized. 

```{r}
# read calls data

# read aggregated calls data
calls = read_csv("aggregated_calls_full_nmcspc.csv") 

# probe_sequence NOT unique
# need to generate unique PROBE_ID
# PROBE_ID in raw_data_complete.csv is paste0(SEQ_ID, ";", POSITION)
calls$PROBE_ID = paste0(calls$seq_id, ";", calls$position)
calls <- calls %>% select(container:position, PROBE_ID, everything()) # rearrange columns

# keep only patients that appear in patient_sample_key
calls <- calls %>% select(container:PROBE_ID, any_of(patient_key$id))

# need calls_long for later 
calls_long <- calls %>%
  select(PROBE_ID, any_of(array_id_key$id)) 

# remove peptides that have zero calls in ALL subjects
calls <- calls[ apply( calls %>% select(any_of(patient_key$id)) , 1, function(x){ !all(x==0) } ) , ]
```

# Reproducibility of Replicates

We have assessed the issue of replicate reproducibility by looking at (Pearson and Spearman's Rank) correlation coefficients between patients' fluorescence levels (both with and without $\log_2$ transformation). Another approach is to measure how much variation the technical replicates are contributing to the overall variation in the data. If the replicates are reproducible, ie. they "largely agree with one another", then the technical variation in the dataset should be minimal.  

Everytime when the fluorescence levels were measured (with replicates) for patient's stage effects, there are two sources of random variation at play, namely  

* patient/subject ``random effect``:  this reflects the biological variation of a patient (as opposed to the ``fixed effect`` term, which would be the cancer stage effect in this experiment)
* (residual) random error: measuring replicates of a patient is itself a source of technical variation.   

Specifically,  

$$y_{ijk} = \mu + \beta_i + b_j + \epsilon_{ijk},$$  

where  

* $y_{ijk}$ denotes the $\log_2$ fluorescence level of a replicate,
* $\mu$ denotes the grand mean/intercept,
* $\beta_i$ denotes the fixed effect term, ie. cancer stage, with $i$ indexing the patients' cancer stage,
* $b_j$ denotes the random effect term, ie. individual patient, with $j$ indexing the patients,
* $\epsilon_{ijk}$ denotes the (residual) random error of the model, with $k$ indexing the replicates. 

This is the linear mixed-effects model, which we deploy using the R package ``lme4``. The model estimates the two sources of variation: $\hat{\sigma}^2_b$ (biological variation) and $\hat{\sigma}^2_\epsilon$ (technical variation). Ideally, biological variation should dominate technical variation since the replicates' variance $\hat{\sigma}^2_{\epsilon}$ should be minimal. Hence, we are interested in the estimated proportion of random-effect variance to total variance  

$$\dfrac{\hat{\sigma}^2_b}{ \hat{\sigma}^2_b + \hat{\sigma}^2_\epsilon},$$  

and we would like to see if this ratio is close to one. For each of the 177,604 peptides, we deploy this mixed-effect model, and plot the histogram of the estimated proportions of variances.

```{r, fig.height = 5}
# get estimated proportion of variances
lmer_var_ratio <- lmer_result[,'variance_id'] / ( lmer_result[,'variance_id'] + lmer_result[,'variance_residual']  )
hist(lmer_var_ratio, breaks = 100, xlab = "estimated proportion of variances",
     main = "Histogram of peptide-level proportion of \nrandom-effect variance to total variance")
```

As expected, the histogram amasses at values near one, indicating that most of the variation in the ($\log_2$) fluorescence data is attributable to the biological variation of the patients and not the technical replicates themselves, which also suggests reproducibility of the replicates. The little spike at zero estimated proportions is due to the `r sum(as.numeric(lmer_result[,'singularity']))` singular cases where the fitted random-effect variance $\hat{\sigma}^2_b$ is close/equal to zero. 


# Tests on Binary Calls

We acknowledge the consensus that the binary calls on a peptide of a patient could be conservative -- out of 177,604 peptides, only `r nrow(calls)` of them have at least one call among all patients. If we collapse all patients' peptides to quantify the number of zero- and one- calls, we obtain the following counts:

```{r, cache = T, dependson= c("raw_data_median", "array_id_key", "calls_long")}
# get median_long 
median_long2 <- raw_data_median %>%
  select(PROBE_ID, any_of(array_id_key$id)) 

# rearrange rows and columns to match calls_long & median_long
calls_long <- calls_long[, match(colnames(median_long2), colnames(calls_long))]
calls_long <- calls_long[ match(median_long2$PROBE_ID, calls_long$PROBE_ID) , ]

# pivot_longer
median_long2 <- median_long2 %>% select(-PROBE_ID) %>%
  pivot_longer(cols = everything(), names_to = "id", values_to = "fluorescence") 
calls_long <- calls_long %>% select(-PROBE_ID) %>%
  pivot_longer(cols = everything(), names_to = "id", values_to = "calls") 

# get the data frame for ggplot
calls_fl_df <- data.frame(
  calls = factor(calls_long$calls),
  fluorescence = median_long2$fluorescence
)

kable(janitor::tabyl(calls_fl_df$calls) %>% janitor::adorn_pct_formatting() %>%
  rename(calls = "calls_fl_df$calls", patient_peptide_counts = n),
       format = "latex", row.names= F, booktabs = T ) %>%
  kable_styling(latex_options = c("hold_position"))
```

To verify that positive calls are associated with stronger signals (remember that ``call`` = 1 if fluorescence levels meet a certain signal threshold in at least two of the replicates), we plot the boxplot of $\log_2$ fluorescence levels for all peptides across all patients, comparing between those that are associated with positive calls and those with zero-calls. 

```{r, fig.height = 3.5, cache = T, dependson=c("calls_fl_df")}
ggplot(calls_fl_df, aes(x = calls, y = fluorescence, fill = calls)) +
  geom_boxplot() +
  labs(title = "Boxplots of Fluorescence Levels per Peptide per Patient", 
       x = "Calls per peptide per patient", y = "Median Fluorescence Levels") +
  theme(panel.background = element_rect(fill = "grey90"),
        panel.grid.major = element_line(color = "white"),
        panel.grid.minor = element_line(color = "white"),
        plot.title = element_text(hjust = 0.5))
```

Even among the `r nrow(calls)` peptides (that have at least one call), most of them have very few positive calls among all patients. We plot the distribution of the sum of calls among all patients for each of these peptides.

```{r, fig.height = 3.8}
# see row sums of calls
calls_rowsums <- calls %>% select(any_of(patient_key$id)) %>% rowSums()
calls_rowsums_df <- as.data.frame(table(calls_rowsums)) %>%
  mutate(Freq = ifelse(as.numeric(calls_rowsums) >=10, sum(Freq[as.numeric(calls_rowsums) >=10]), Freq))  %>%
  filter(as.numeric(calls_rowsums) <=10) %>%
  mutate(calls_rowsums = ifelse(as.numeric(calls_rowsums) == 10 , "10 and above", calls_rowsums),
         calls_rowsums = factor(calls_rowsums, levels = c(1:9,"10 and above")))
ggplot(data=calls_rowsums_df, aes(x=calls_rowsums, y=Freq)) +
  geom_bar(stat="identity", fill="steelblue")+
  geom_text(aes(label=paste0(Freq)), vjust=-.3, size=3)+
  labs(x = "Sum of calls among all patients for a peptide",
       title = paste0("Sum of Calls Among All Patients for Each Peptide (Total ", nrow(calls), " Peptides)")) +
  theme(plot.title = element_text(hjust = 0.5))
```

For example, there are `r calls_rowsums_df$Freq[1] ` peptides with only one positive call among all `r nrow(patient_key)` patients. For each of these `r nrow(calls)` peptides, we run a logistic regression based on these binary calls of the patients in order **to determine if calls are significantly different among patients of different cancer stages**.  

Specifically, for each of these `r nrow(calls)` peptides, we fit the following model:

$$
\text{logit} \left( y^{\text{calls}}_{ij} \right)
= \mu + \beta_i + \epsilon_{ij},
$$

where  

* $y^{\text{calls}}_{ij}$ denotes the binary call of the peptide of a patient: 1 if the fluorescence levels meet the signal threshold in at least two replicates of the patient, and 0 otherwise,
* $\mu$ denotes the grand mean/intercept,
* $\beta_i$ denotes the cancer stage,
* $\epsilon_{ij}$ denotes the random error of the model, with $j$ indexing the patients,  

and compute the deviance p-values: (null_deviance - residual_deviance) $\sim$ $\chi^2$ with `r length(unique(patient_key$stage)) - 1` degrees of freedom. We plot the histogram of the `r nrow(calls)` p-values.

```{r}
# plot histogram of p-values
hist(logreg_pval, breaks = 50, freq = T, main = "Logistic Regression Deviance p-values", xlab = "p-values ")
```

It appears that there are hardly any signals of different calls pattern among patients of different cancer stages, which corroborates with the results in Figures 2, 3 and 4 in the main manuscript. After correcting for false discovery rate, no peptides appear to be significant. 

# Tests on Fluorescence Levels 

We lose a lot of information due to the conservative signal threshold of the calls data. Instead, in this section, we will investigate antibody response level among patients of different cancer stages using the fluorescence data. In the next section, we will utilize both fluorescence and calls data to investigate our hypothesis and compare our results. First, we perform one-way ANOVA (analysis of variance) to test the following:  
$H_0$: Antibody responses are the same for patients at different stages of prostate cancer.  
$H_1$: Antibody responses are not the same for patients at different stages of prostate cancer. 

After getting p-values for all the peptides, we plot the p-value histogram.

```{r}
# get p-values histogram and FDR 
all_anova <- anova_func(all_anova_pval)
```

If cancer-stage effect is not present in our peptide array data, then the p-values from the ANOVA would have a uniform distribution between 0 and 1, and we expect to see a rather flat-shaped histogram of p-values.  
  
However, the p-values histogram exhibits large counts of significant p-values (p-values close to zero), and the shape of histogram flattens off exponentially with larger p-values. Such a large count of significant p-values may not be explained by false discovery alone, and that perhaps cancer-stage effect is indeed present in some of the peptides in our profile. The red-shaded regions of the histogram represents the estimated proportion of non-null peptides in the dataset based on Storey's q-values calculation obtained via the R package ``fdrtool``. The q-value is similar to the well known p-value, except that it is a measure of significance in terms of the false discovery rate rather than the false positive rate.  

We apply the Benjamini-Hochberg (BH) method on the ANOVA p-values to control for false discovery rate (FDR). The peptide counts at various BH FDR thresholds are tabulated below. 

```{r}
kable( count.func(all_anova$anova_BH, seq(0.01, 0.1, by = 0.01)),
       format = "latex", row.names= NA, col.names = NA, booktabs = T )%>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%
  row_spec(1, hline_after = T)
```

Here are also the peptide counts at different q-values thresholds.

```{r}
kable( count.func(all_anova$anova_qval, seq(0.01, 0.1, by = 0.01)),
       format = "latex", row.names= NA, col.names = NA, booktabs = T )%>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%
  row_spec(1, hline_after = T)
```

It appears that the BH method gives a more conservative significant peptide counts.  

## Visualizing Results

```{r}
BH_FDR_counts <- length(all_anova$anova_BH[all_anova$anova_BH <= 0.05])
```

It appears that we have identify `r BH_FDR_counts` significant peptides at 5\% BH FDR. We would like to obtain some graphical representations to illustrate how the $\log_2$ fluorescence levels differ across different cancer stages for these peptides.  

For each peptide, we remove the grand mean (row mean) of the $\log_2$ fluorescence levels for all patients, before applying the following visualization techniques:  

* Principal Component Analysis (PCA)
* t-distributed Stochastic Neighbor Embedding (t-SNE)
* Heatmap  

Based on our one-way-ANOVA model assumption, if there is no cancer-stage effect, we expect these residual $\log_2$ fluorescence to be random noises. Any observed (clustering) patterns among these residual data points reveal the effects of various stages of prostate cancer.  

For purpose of uniformity, we also use the same color scheme to distinguish the different stages of cancer patients (notice how the spectrum of colors changes with severity of the cancer stages):  

* navy for healthy subjects
* cornflower_blue for ``new_dx`` newly diagnosed patients
* turquoise for ``nmCSPC`` patients
* light_pink for ``mCSPC`` patients -- these patients have no technical replicates and are excluded from this analysis
* dark_orange for ``nmCSPC`` patients
* dark_red for ``mCSPC`` patients 

```{r}
BH_FDR_cutoff <- 0.05

# all_anova
anova_dat <- raw_data_median %>% 
  select(PROBE_ID, any_of(patient_key$id)) %>% 
  mutate(anova_BH = all_anova$anova_BH[raw_data_median$PROBE_ID]) %>%
  filter(anova_BH <= BH_FDR_cutoff) %>%
  select(-anova_BH)

anova_dat_demean <- sweep(as.matrix(anova_dat %>% select(-PROBE_ID)), 1, 
                          rowMeans(as.matrix(anova_dat %>% select(-PROBE_ID))), "-") # centering by row

# make sure stage aligns with anova_dat_demean
visual_iii <- match( colnames(anova_dat_demean) , patient_key$id )
visual_stage <- patient_key$stage[visual_iii]

# colors and shapes for the visualization techniques
cols = pal[ match(visual_stage, names(pal)) ]
shapes = shp[  match(visual_stage, names(shp)) ]
```

First, the PCA plot is given by

```{r}
# PCA

# svd
sv.dat <- sweep(t(anova_dat_demean), 2, colMeans(t(anova_dat_demean)), "-") # centering
sv <- svd(sv.dat)
V <- sv$v
D <- sv$d
U <- sv$u

# variance explained
pca.var <- D^2/sum(D^2) 
pca.cumvar <- cumsum(pca.var)

# plot PCA
par(mfrow = c(1,2), pty = "s", mar = c(2.2,2.3,1.5,0.45), mgp = c(1.4,0.4,0),
    cex.axis = 0.84, cex.lab = 0.84, cex.main = 0.84, tcl = -0.4)
PCload.func(cols, shapes, U, D, 1, 2, pca.var, title = "PC2 vs PC1") # PC loadings (PC2 vs PC1)
legend('topright', pch = shp, col = pal, cex = 0.5,
       c("normal",  "new_dx", "nmCSPC", "mCSPC", "nmCRPC", "mCRPC") )
PCload.func(cols, shapes, U, D, 3, 2, pca.var, title = "PC2 vs PC3") # PC loadings (PC2 vs PC3)
```

From the "\textit{PC2 vs PC1}" plot, we observe that all mCRPC points are clustered at the topright of the panel, whereas nmCRPC observations hover at the bottom of panel. The percentage of variance explained for each principal component (PC) is shown on the axis. Note that the first principal component manages to capture most of the variation in the data.   

Next, we obtain the t-SNE plot. Just like PCA, t-SNE is a dimensionality reduction technique which was first introduced by van der Maaten and Hinton in 2008.  

```{r, fig.height=6, cache = T, dependson=c("anova_dat_demean")}
# t-SNE

# how to specify parameter
# refer: https://lvdmaaten.github.io/tsne/User_guide.pdf
tsnedat <- unname(t(anova_dat_demean)) # dim(X) = N samples by D dimensions 
initdim <- 90 
perplex <- 30 

set.seed(10)
tsne_anova <- Rtsne(tsnedat, initial_dims = initdim, perplexity = perplex,
                    theta = 0, check_duplicates = F, max_iter = 50000L, eta = 50)

# t-SNE plot
plot(tsne_anova$Y, ylab = "", xlab = "", col = cols, pch = shapes, main = "t-SNE plot")
legend('topright', pch = shp, col = pal,
       c("normal",  "new_dx", "nmCSPC", "mCSPC", "nmCRPC", "mCRPC") )
```

From the t-SNE plot, we observe that:  

* The mCRPC points are clustered together near top-left of the plot.
* Most of the nmCRPC patients are not too far off from the mCRPC subjects. There seems to be 2 clusters of nmCRPC patients.
* Most of the new_dx and nmCSPC patients are clustered at the bottom left and middle of the plot.
* Normal subjects are somewhat "all over the place".  

Finally, we plot the heatmap of the $\log_2$ fluorescence (after removing row means) of these peptides at 5\% BH-FDR.

```{r, fig.height = 10, fig.width = 8, cache = T, dependson=c("anova_dat_demean")}
# heatmap color palette
cls <- colorRampPalette(c("navy", "honeydew", "firebrick3", "brown"))(n = 1024)

heatmap3(anova_dat_demean, 
         col = cls, # specify colors 
         ColSideColors = cols, # specify patient color code
         labCol = visual_stage, # specify patient
         ColSideLabs = "stages", 
         labRow = "",
         xlab = "Patients",
         # legendfun=function() showLegend(col = c("navy", "cornflowerblue", "turquoise1", "orchid1", 
         #                                         "darkorange1", "firebrick1"),
         #                                 legend = c("normal",  "new_dx", "nmCSPC", "mCSPC", "nmCRPC", "mCRPC"),
         #                                 cex = 1.2,
         #                                 lwd = 5  )
)
```

Again, we observe similar patterns that nmCSPC (colored turquoise) and new_dx (colored cornflowerblue) subjects are mostly clustered to the left part of the heatmap, whereas the mCRPC (colored red) patients and most of the nmCRPC patients (colored darkorange) are clustered to the right part of the heatmap. Again, normal subjects (colored navy) are "all over the place".

## Gene-Set Analysis

```{r}
# read uniprot_gene csv
uniprot_gene <- read_csv("uniprot_data_entrez.csv", col_types = cols_only(
  seq_id = col_character(),
  uniprot_id = col_character(),
  gene_symbol = col_character(),
  entrez_gene_id_pete = col_double(),
  gene_names = col_character(),
  protein_names = col_character()
)) %>% filter(!(is.na(seq_id)) & !(is.na(entrez_gene_id_pete))) %>%
  select(seq_id, uniprot_id, gene_symbol, entrez_gene_id_pete, gene_names, protein_names)

## manually changing entrez_gene_id for PCA10 & PRO29
uniprot_gene$entrez_gene_id_pete[uniprot_gene$seq_id == "PCA10"] <- 28912
uniprot_gene$entrez_gene_id_pete[uniprot_gene$seq_id == "PRO29"] <- NA
uniprot_gene <- uniprot_gene[!(is.na(uniprot_gene$entrez_gene_id_pete)),]

# get unique seq_id that are associated with significant peptides
signif_seq_id <- unique( raw_data_median$SEQ_ID[raw_data_median$PROBE_ID %in% anova_dat$PROBE_ID] )
signif_seq_id <- signif_seq_id[!(is.na(signif_seq_id))] # just in case

# convert these into binary vector
seq_id_ok <- as.numeric(uniprot_gene$seq_id %in% signif_seq_id)
names(seq_id_ok) <- uniprot_gene$entrez_gene_id_pete
```

Recall that the 177604 peptides correspond to 1611 proteins, and `r nrow(uniprot_gene)` of these proteins have matching genes in *uniprot_gene_entrez.csv*. The `r BH_FDR_counts` significant peptides at 5\% BH-FDR are associated with `r sum(seq_id_ok)` proteins with matching genes. In this analysis, we deem a protein to be significant if it has at least one significant associated peptide. We shall utilize this information to perform gene-set enrichment analysis.  

Specifically, we investigate if there are any pre-specified gene-sets that are enriched for the genes associated with the list of significant peptides. These pre-determined gene-sets are defined based on their functional categories or biological properties, such as the Gene Ontology (GO) annotations. Enriched gene-sets could reflect the biological signals in the peptide microarray data.  
The gene-set-analysis is performed with the R package ``allez``. We shall consider gene-sets containing at least 2 genes with Bonferroni-corrected enrichment p-values not exceeding 5\% . 

```{r, cache = T, dependson=c("seq_id_ok"), echo = T}
# gene-set analysis via allez!
allez.go <- allez(seq_id_ok, lib = "org.Hs.eg", idtype = "ENTREZID", sets = "GO")

# nominal alpha for enrichment score
nom.alpha <- 0.05 

# minimum number of genes in gene-set to be considered 
min.num.gene <- 2
```

We display the waterfall plot of the gene-set-analysis results. It turns out that in this case, there is only one significantly enriched gene set, since most of the proteins considered in this case are associated with at least one significant peptide.

```{r, fig.height = 2.3, fig.width=9, echo = T}
# Waterfall plot
allezPlot(allez.go, n.cell = min.num.gene, nominal.alpha = nom.alpha)
```


## Comparing Healthy Subjects vs Cancer Patients

The one-way ANOVA is helpful in revealing peptides that exhibit significant difference in the group means of $\log_2$ fluorescence levels among patients at different stages of cancer. We may be interested to find out, for example, **if antibody response is significantly higher in cancer patients than in healthy subjects**. Using the linear-contrast method, we are able to further test this hypothesis after computing the ANOVA.  

Specifically, it is a one-sided t-test comparing the averages of the $\log_2$ fluorescence levels among cancer patients and among healthy subjects, except that now, the degrees of freedom and the standard error of the t-test statistic are based on the one-way ANOVA computed previously. This allows us to pull samples from different cancer stages and increase power -- with the caveat/assumption that the patients of different cancer stages share the same level of variation in antibody response. (Note that we could avoid doing the multiplicity control like Tukey's Honest Significant Difference if we are only interested in a particular contrast/comparison and not every possible combination of pairwise comparisons)

```{r}
# again, make sure stage aligns with anova_dat
contr_iii <- match( colnames(anova_dat %>% select(-PROBE_ID)) , patient_key$id )
contr_stage <- patient_key$stage[contr_iii]

# get group means and MSE
BH_FDR_normal <- apply(as.matrix(anova_dat %>% select(-PROBE_ID)), 1, 
                       function(x){mean(x[contr_stage == "normal"])})
BH_FDR_cancer <- apply(as.matrix(anova_dat %>% select(-PROBE_ID)), 1, 
                       function(x){mean(x[contr_stage != "normal"])})
BH_FDR_MSE <- apply(as.matrix(anova_dat %>% select(-PROBE_ID)), 1, function(x){
  fit1 <- lm(x ~ contr_stage)
  return(deviance(fit1)/df.residual(fit1))
})

# get linear contrast p-values
normal_counts <- length(contr_stage[contr_stage == "normal"])
cancer_counts <- length(contr_stage[contr_stage != "normal"])
BH_FDR_tstat <- (BH_FDR_cancer - BH_FDR_normal) / sqrt(BH_FDR_MSE * (1/cancer_counts + 1/normal_counts))
BH_FDR_tstat_pval <- 1 - pt( BH_FDR_tstat, df = (length(unique(patient_key$stage)) - 1) )
```

There are `r length(BH_FDR_tstat_pval[BH_FDR_tstat_pval <= 0.05])` peptides that are associated with higher antibody response in cancer patients than in normal subjects. We tabulate these peptides with their differences (of average $\log_2$ fluorescence levels between cancer and healthy subjects) and p-values of their t-statistics. 

```{r}
# output these probes and their t-stat p-val
contr_df <- data.frame( peptide_id = anova_dat$PROBE_ID[BH_FDR_tstat_pval <= 0.05],
                        difference = round((BH_FDR_cancer - BH_FDR_normal)[BH_FDR_tstat_pval <= 0.05],4),
                        tstat_pvalues = round(BH_FDR_tstat_pval[BH_FDR_tstat_pval <= 0.05],4))
contr_df <- contr_df[base::order(contr_df$tstat_pvalues),]

kable( contr_df, format = "latex", row.names= F, booktabs = T, longtable = T )%>%
  kable_styling(font_size = 5, latex_options = c("hold_position", "repeat_header")) 
```

# Tests on Fluorescence Levels with Binary-calls-filtering

```{r}
names(calls_rowsums) <- calls$PROBE_ID
BH_FDR_calls_counts <- sum(as.numeric(names(all_anova$anova_BH[all_anova$anova_BH<=0.05]) %in% 
                                        names(calls_rowsums[calls_rowsums >=1])))
```

Previously we identified `r BH_FDR_counts` significant peptides at 5\% BH FDR. Out of these peptides, `r BH_FDR_calls_counts` of them also have at least one call among all patients. We will use these calls as filtering criterion on the significant peptides -- in this section, we will examine whether zooming in on these `r BH_FDR_calls_counts` significant peptides help to improve clustering patterns that we would observe in PCA, t-SNE and heatmap. 

## Visualization based on Filtered Significant Peptides

```{r}
anova_dat <- raw_data_median %>% 
  select(PROBE_ID, any_of(patient_key$id)) %>% 
  mutate(anova_BH = all_anova$anova_BH[raw_data_median$PROBE_ID]) %>%
  filter(anova_BH <= BH_FDR_cutoff) %>%
  filter(PROBE_ID %in% calls$PROBE_ID) %>%
  select(-anova_BH)


anova_dat_demean <- sweep(as.matrix(anova_dat %>% select(-PROBE_ID)), 1, 
                          rowMeans(as.matrix(anova_dat %>% select(-PROBE_ID))), "-") # centering by row

# make sure stage aligns with anova_dat_demean
visual_iii <- match( colnames(anova_dat_demean) , patient_key$id )
visual_stage <- patient_key$stage[visual_iii]

# colors and shapes for the visualization techniques
cols = pal[ match(visual_stage, names(pal)) ]
shapes = shp[  match(visual_stage, names(shp)) ]
```

Based on the `r BH_FDR_calls_counts` significant peptides, the PCA plots become

```{r}
# PCA

# svd
sv.dat <- sweep(t(anova_dat_demean), 2, colMeans(t(anova_dat_demean)), "-") # centering
sv <- svd(sv.dat)
V <- sv$v
D <- sv$d
U <- sv$u

# variance explained
pca.var <- D^2/sum(D^2) 
pca.cumvar <- cumsum(pca.var)

# plot PCA
par(mfrow = c(1,2), pty = "s", mar = c(2.2,2.3,1.5,0.45), mgp = c(1.4,0.4,0),
    cex.axis = 0.84, cex.lab = 0.84, cex.main = 0.84, tcl = -0.4)
PCload.func(cols, shapes, U, D, 1, 2, pca.var, title = "PC2 vs PC1") # PC loadings (PC2 vs PC1)
legend('topright', pch = shp, col = pal, cex = 0.5,
       c("normal",  "new_dx", "nmCSPC", "mCSPC", "nmCRPC", "mCRPC") )
PCload.func(cols, shapes, U, D, 3, 2, pca.var, title = "PC2 vs PC3") # PC loadings (PC2 vs PC3)
```

The clustering patterns here is similar to the pattern we observe in the PCA plot for all `r BH_FDR_counts` peptides. The t-SNE plot now becomes

```{r, fig.height=6, cache = T, dependson=c("anova_dat_demean")}
tsnedat <- unname(t(anova_dat_demean)) # dim(X) = N samples by D dimensions 
initdim <- 90 
perplex <- 30 

set.seed(10)
tsne_anova <- Rtsne(tsnedat, initial_dims = initdim, perplexity = perplex,
                    theta = 0, check_duplicates = F, max_iter = 50000L, eta = 50)

# t-SNE plot
plot(tsne_anova$Y, ylab = "", xlab = "", col = cols, pch = shapes, main = "t-SNE plot")
legend('topleft', pch = shp, col = pal,
       c("normal",  "new_dx", "nmCSPC", "mCSPC", "nmCRPC", "mCRPC") )
```

Again, clustering pattern is somewhat similar. Finally, the heatmap becomes

```{r, fig.height = 10, fig.width = 8, cache = T, dependson=c("anova_dat_demean")}
# heatmap color palette
cls <- colorRampPalette(c("navy", "honeydew", "firebrick3", "brown"))(n = 1024)

heatmap3(anova_dat_demean, 
         col = cls, # specify colors 
         ColSideColors = cols, # specify patient color code
         labCol = visual_stage, # specify patient
         ColSideLabs = "stages", 
         labRow = "",
         xlab = "Patients",
         # legendfun=function() showLegend(col = c("navy", "cornflowerblue", "turquoise1", "orchid1", 
         #                                         "darkorange1", "firebrick1"),
         #                                 legend = c("normal",  "new_dx", "nmCSPC", "mCSPC", "nmCRPC", "mCRPC"),
         #                                 cex = 1.2,
         #                                 lwd = 5  )
)
```

The clustering pattern that reveals cancer stage effect is more obvious in the previous section (when all `r BH_FDR_counts` peptides are considered) than what we observe here. 

## Gene Set Analysis

```{r}
# read uniprot_gene csv
uniprot_gene <- read_csv("uniprot_data_entrez.csv", col_types = cols_only(
  seq_id = col_character(),
  uniprot_id = col_character(),
  gene_symbol = col_character(),
  entrez_gene_id_pete = col_double(),
  gene_names = col_character(),
  protein_names = col_character()
)) %>% filter(!(is.na(seq_id)) & !(is.na(entrez_gene_id_pete))) %>%
  select(seq_id, uniprot_id, gene_symbol, entrez_gene_id_pete, gene_names, protein_names)

## manually changing entrez_gene_id for PCA10 & PRO29
uniprot_gene$entrez_gene_id_pete[uniprot_gene$seq_id == "PCA10"] <- 28912
uniprot_gene$entrez_gene_id_pete[uniprot_gene$seq_id == "PRO29"] <- NA
uniprot_gene <- uniprot_gene[!(is.na(uniprot_gene$entrez_gene_id_pete)),]

# CAREFUL!
# further filter uniprot_gene to limit them to proteins associated with filtered peptides based on calls
uniprot_gene  <- uniprot_gene[ uniprot_gene$seq_id %in% (unique(calls$seq_id)) ,]

# get unique seq_id that are associated with significant peptides
signif_seq_id <- unique( raw_data_median$SEQ_ID[raw_data_median$PROBE_ID %in% anova_dat$PROBE_ID] )
signif_seq_id <- signif_seq_id[!(is.na(signif_seq_id))] # just in case

# convert these into binary vector
seq_id_ok <- as.numeric(uniprot_gene$seq_id %in% signif_seq_id)
names(seq_id_ok) <- uniprot_gene$entrez_gene_id_pete
```

We repeat our gene set enrichment analysis, but only with the proteins that are associated with peptides with at least one call among all patients. Hence, in this case, we only consider `r nrow(uniprot_gene)` proteins with matching genes, and `r sum(seq_id_ok)` of them are associated with significant peptides with at least one call among all patients. We use this information to perform our gene set analysis. 

```{r, cache = T, dependson=c("seq_id_ok"), echo=T}
# gene-set analysis via allez!
allez.go <- allez(seq_id_ok, lib = "org.Hs.eg", idtype = "ENTREZID", sets = "GO")

# nominal alpha for enrichment score
nom.alpha <- 0.05 

# minimum number of genes in gene-set to be considered 
min.num.gene <- 2
```

We display the waterfall plot of the gene-set-analysis results. 

```{r, echo = T, fig.width=9, fig.height=5.3}
# Waterfall plot
allezPlot(allez.go, n.cell = min.num.gene, nominal.alpha = nom.alpha)
```

The waterfall plot was constructed by finding the significant (Bonferroni-corrected p-value < 0.05) GO term having the largest overlap with genes associated with proteins that have at least one significant peptide with at least one call among all patients (nucleic acid binding GO:0003676) and placing it in the top row of the figure. We next removed these genes from the list and found the significant GO term having the highest overlap with the remainder of the list (non-membrane-bounded organelle GO: 0043228). This process is repeated, and genes identified by this sequential process are counted along the x-axis, and the overlap between the GO terms can be visually assessed. Shading under the 'waterfall' component of the graph indicates genes that were annotated to previously named categories.  

We also tabulate the enriched/overrepresented GO terms. The last column of the table shows the genes associated with proteins that have at least one significant peptide with at least one call among all patients.

```{r}
# tabulate enriched gene-set with signif genes only
allez.tab <- allezTable(allez.go, symbol = T, n.cell = min.num.gene, nominal.alpha = nom.alpha)
rich.vec <- NULL
rich.count.vec <- NULL
for (i in 1 : dim(allez.tab)[1])
{
  check.string <- strsplit(allez.tab[i,6], split = ";")[[1]]
  ori.count <- length(check.string)
  rich.ok <- check.string %in% (uniprot_gene$gene_names)[match(signif_seq_id, uniprot_gene$seq_id)]
  rich.count <- length(rich.ok[rich.ok])
  rich.vec <- c( rich.vec, paste(check.string[rich.ok],collapse="; ") )
  rich.count.vec <- c( rich.count.vec, paste(rich.count,"/",ori.count, sep = "") )
}
allez.tab2 <- cbind(allez.tab[,1:3], rich.count.vec, allez.tab[,5], rich.vec)
allez.tab2[,5] <- round( as.numeric(allez.tab2[,5]), 4)
colnames(allez.tab2) <- colnames(allez.tab)

kable(allez.tab2, format = "latex", digits = 4, booktabs = T, longtable = T) %>%
      kable_styling(latex_options = c("hold_position", "repeat_header"),
                    full_width = F, font_size = 6) %>%
      column_spec(2, width = "12em") %>%
      column_spec(7, width = "16em")
```

We could compare these enriched gene-sets with the results in Figures 4 and 5 of the main manuscript.  

## Comparing Healthy Subjects vs Cancer Patients

Here, we repeat the linear contrast method on these `r BH_FDR_calls_counts` peptides to determine if fluorescence levels of these peptides are significantly higher among cancer patients than among healthy subjects.

```{r}
# again, make sure stage aligns with anova_dat
contr_iii <- match( colnames(anova_dat %>% select(-PROBE_ID)) , patient_key$id )
contr_stage <- patient_key$stage[contr_iii]

# get group means and MSE
BH_FDR_normal <- apply(as.matrix(anova_dat %>% select(-PROBE_ID)), 1, 
                       function(x){mean(x[contr_stage == "normal"])})
BH_FDR_cancer <- apply(as.matrix(anova_dat %>% select(-PROBE_ID)), 1, 
                       function(x){mean(x[contr_stage != "normal"])})
BH_FDR_MSE <- apply(as.matrix(anova_dat %>% select(-PROBE_ID)), 1, function(x){
  fit1 <- lm(x ~ contr_stage)
  return(deviance(fit1)/df.residual(fit1))
})

# get linear contrast p-values
normal_counts <- length(contr_stage[contr_stage == "normal"])
cancer_counts <- length(contr_stage[contr_stage != "normal"])
BH_FDR_tstat <- (BH_FDR_cancer - BH_FDR_normal) / sqrt(BH_FDR_MSE * (1/cancer_counts + 1/normal_counts))
BH_FDR_tstat_pval <- 1 - pt( BH_FDR_tstat, df = (length(unique(patient_key$stage)) - 1) )
```

There are `r length(BH_FDR_tstat_pval[BH_FDR_tstat_pval <= 0.05])` peptides that are associated with higher antibody response in cancer patients than in normal subjects. We tabulate these peptides with their differences (of average $\log_2$ fluorescence levels between cancer and healthy subjects) and p-values of their t-statistics. 

```{r}
# output these probes and their t-stat p-val
contr_df <- data.frame( peptide_id = anova_dat$PROBE_ID[BH_FDR_tstat_pval <= 0.05],
                        difference = round((BH_FDR_cancer - BH_FDR_normal)[BH_FDR_tstat_pval <= 0.05],4),
                        tstat_pvalues = round(BH_FDR_tstat_pval[BH_FDR_tstat_pval <= 0.05],4))
contr_df <- contr_df[base::order(contr_df$tstat_pvalues),]

kable( contr_df, format = "latex", row.names= F, booktabs = T, longtable = T )%>%
  kable_styling(font_size = 6, latex_options = c("hold_position", "repeat_header")) 
```

Another form of verification using the calls data would be as follows: we check the sum of positive calls for each cancer stage for these `r length(BH_FDR_tstat_pval[BH_FDR_tstat_pval <= 0.05])` peptides. 

```{r}
kable( calls[calls$PROBE_ID %in% anova_dat$PROBE_ID[BH_FDR_tstat_pval <= 0.05], ] %>%
  select(any_of(patient_key$id)) %>% 
  colSums() %>% enframe(name = "id", value = "sum_of_calls") %>%
  inner_join(patient_key) %>%
  group_by(stage) %>% 
  summarize( n = n(),
             total_calls = sum(sum_of_calls) ),
  format = "latex", row.names= F, booktabs = T) %>%
  kable_styling(latex_options = c("hold_position")) 
```

As expected, almost all of the positive calls for these peptides come from cancer patients instead of normal subjects. For purpose of visualization, we also produce the boxplots of $\log_2$ fluorescence levels of subjects from these two groups for some of these peptides.

```{r, fig.height=2.9}
# boxplot specs
stage2 <- as.character(contr_stage)
stage2[stage2 != "normal"] <- "cancer"
stage2 <- factor(stage2)
boxplot_col <- c("red", "blue")
boxplot_func <- function(mat, col = boxplot_col, draw){
  # par(mar = c(2, 4.5, 2.3, 1),cex = 0.84)
  graphics::boxplot( as.numeric(mat[draw,]) ~ stage2, 
                     col=col, horizontal=TRUE, las=1, xlab = "log2(fluorescence)", ylab = "",
                     main= paste(row.names(mat)[draw]) )
}

# get boxplot matrix ready
boxplot_mat <- anova_dat[BH_FDR_tstat_pval <= 0.05,]
boxplot_mat <- as.matrix(boxplot_mat %>% select(-PROBE_ID))
row.names(boxplot_mat) <- anova_dat$PROBE_ID[BH_FDR_tstat_pval <= 0.05]
boxplot_mat <- boxplot_mat[order(BH_FDR_tstat_pval[BH_FDR_tstat_pval <= 0.05]),]

# boxplot
boxplot_func(mat = boxplot_mat, draw = 1)
boxplot_func(mat = boxplot_mat, draw = 2)
boxplot_func(mat = boxplot_mat, draw = 3)
```

# Concluding Remarks

* Calls are conservative -- most of the calls are zero anyway. No cancer-stage signals are observable from the calls data.
* There appears to be cancer stage effects present in the $\log_2$ fluorescence data with or without filtering peptides based on calls. The clustering patterns of data points based on cancer stages seem to be slightly more obvious without calls-filtering.  
